Simple Type Theory is not too Simple

### 2021-04-09 article.tex
Remove Lamport reference and use Coquelicot instead

Last, it seems that our experience in Isabelle to formalize schemes was much smoother than the corresponding one in Lean, since Kevin Buzzard reported on his blog the serious difficulties he faced during his formalization in Lean:
\begin{quote}
	The project is completely incompatible with modern Lean and mathlib, but if you compile it then you get a sorry-free proof that an affine scheme is a scheme. During our proof we ran into a huge problem because our blueprint, the Stacks Project, assumed that $R[1/f][1/g]=R[1/fg]$, and this turns out to be unprovable for Lean’s version of equality [\dots]. The Lean community wrestled with this idea, and has ultimately come up with a beefed-up notion of equality for which the identity is now true.\footnote{\url{https://xenaproject.wordpress.com/2020/06/05/the-sphere-eversion-project/}}
\end{quote}	
As we understand it, this difficulty basically comes from defining the sheaf structure of $\spec$ first on the basis given by the so-called \textit{basic open sets} before extending it to all open subsets. We did not meet this difficulty in Isabelle, since we did not follow this sheaf-on-a-basis approach and the sheaves in our library are always defined from the get-go on all open subsets instead. \\
Also, Buzzard \textit{et al.} mention another difficulty encountered when proving that an affine scheme is a scheme. 
\begin{quote}
	This means that rewriting the equality $\iota(U) = U$ can cause technical problems with data-carrying types such as $\mathscr{O}_X(\iota(U))$ which depends on $\iota(U)$ (although in our case they were surmountable). This is a technical issue with dependent type theory and can sometimes indicate that one is working with the wrong definitions. \cite[3.5]{schemesinLean}
\end{quote}	
where $\iota: \spec \rightarrow \spec$ is here simply the identity map. If, again, we gave a slightly different formulation for the definition of a scheme, following Hartshorne~\cite{hartshorne} instead of the Stacks project \cite{stacksproject} as they did, it seems that this difficulty cannot possibly have a direct counterpart in Isabelle, since it arose from the difference between the so-called \textit{equality types}, also known as \textit{identity types}, and the \textit{definitional equality}, a difference which is peculiar to dependent type theories. \\
If we were much more fortunate with our formalization in Isabelle, it is worth mentioning that this difficulty was eventually overcome in Lean using a trick.  
\begin{quote}
	Fortunately, in our case, Reid Barton pointed out the following extraordinary trick to us: we can define the map $\mathscr{O}_X(\iota(U)) \rightarrow \mathscr{O}_X(U)$ using restriction rather than trying to force it to be the identity! \cite[3.5.]{schemesinLean}
\end{quote}	    
However, all these difficulties in Lean, where three attempts at defining schemes were required, raise the question of whether dependent types could be, at least in some situations, more of a burden than an advantage \cite{PaulsonLamport}. Moreover, in the age of machine learning, one may expect there is a trade-off between the expressiveness of a type theory and one's ability to subject it to automation, for instance through SMT solvers and the so-called hammers embedded in some proof assistants. A dual question is to know how deep one can go formalizing mathematics in a simple type theory. This work provides a first hint at an answer that should not be ruled out: simple type theory is not too simple.

>>>>

Last, it seems that our experience in Isabelle to formalize schemes was much smoother than the corresponding one in Lean, since Kevin Buzzard reported on his blog the serious difficulties he faced during his formalization in Lean:
\begin{quote}
	The project is completely incompatible with modern Lean and mathlib, but if you compile it then you get a sorry-free proof that an affine scheme is a scheme. During our proof we ran into a huge problem because our blueprint, the Stacks Project, assumed that $R[1/f][1/g]=R[1/fg]$, and this turns out to be unprovable for Lean’s version of equality [\dots]. The Lean community wrestled with this idea, and has ultimately come up with a beefed-up notion of equality for which the identity is now true.\footnote{\url{https://xenaproject.wordpress.com/2020/06/05/the-sphere-eversion-project/}}
\end{quote}	
As we understand it, this difficulty basically comes from defining the sheaf structure of $\spec$ first on the basis given by the so-called \textit{basic open sets} before extending it to all open subsets. We did not meet this difficulty in Isabelle, since we did not follow this sheaf-on-a-basis approach and the sheaves in our library are always defined from the get-go on all open subsets instead. \\
Also, Buzzard \textit{et al.} mention another difficulty encountered when proving that an affine scheme is a scheme. 
\begin{quote}
	This means that rewriting the equality $\iota(U) = U$ can cause technical problems with data-carrying types such as $\mathscr{O}_X(\iota(U))$ which depends on $\iota(U)$ (although in our case they were surmountable). This is a technical issue with dependent type theory and can sometimes indicate that one is working with the wrong definitions. \cite[3.5]{schemesinLean}
\end{quote}	
where $\iota: \spec \rightarrow \spec$ is here simply the identity map. If, again, we gave a slightly different formulation for the definition of a scheme, following Hartshorne~\cite{hartshorne} instead of the Stacks project \cite{stacksproject} as they did, it seems that this difficulty cannot possibly have a direct counterpart in Isabelle, since it arose from the difference between the so-called \textit{equality types}, also known as \textit{identity types}, and the \textit{definitional equality}, a difference which is peculiar to dependent type theories. \\
If we were much more fortunate with our formalization in Isabelle, it is worth mentioning that this difficulty was eventually overcome in Lean using a trick.  
\begin{quote}
	Fortunately, in our case, Reid Barton pointed out the following extraordinary trick to us: we can define the map $\mathscr{O}_X(\iota(U)) \rightarrow \mathscr{O}_X(U)$ using restriction rather than trying to force it to be the identity! \cite[3.5.]{schemesinLean}
\end{quote}	    
However, all these difficulties in Lean, where three attempts at defining schemes were required, raise the question of whether dependent types could be, at least in some situations, more of a burden than an advantage. The issue with dependent types has also been highlighted in the context of formalising analysis in the proof assistant Coq.
\begin{quote}
	A more salient issue is the overall use of dependent types for defining some notions. For instance, [\dots] rewriting an expression such as $(f+~g)'(x)$ into $f'(x)+g'(x)$, while straightforward in mathematics, is next to impossible to achieve in Coq, as the proof terms have to be manually built by the user beforehand. Third, these proof terms can quickly blow up, and thus become impossible for the user to exhibit, \textit{e.g.} in the case of iterated derivatives or differentiation under an integral. By the way, integrals in Coq suffer from the same issues. \cite[1.]{boldo2015coquelicot}
\end{quote}	 
Moreover, in the age of machine learning, one may expect there is a trade-off between the expressiveness of a type theory and one's ability to subject it to automation, for instance through SMT solvers and the so-called hammers embedded in some proof assistants. A dual question is to know how deep one can go formalizing mathematics in a simple type theory. This work provides a first hint at an answer that should not be ruled out: simple type theory is not too simple.
